---
title: Multilevel models with {brms}
author: Shane A. Scaggs
date: '2021-12-10'
slug: building-multilevel-models-with-brms
categories:
  - Statistics
tags:
  - Bayes
  - Modeling
  - Data Analysis
---

Multilevel models are a kind of statistical model that estimate parameters at more than one level. They are very useful kind of model because they are suited to the hierarchical and nested organization of the world around us.  In this post, I will demonstrate how to build, describe, and compare multilevel models using tools from `brms`. 

```{r, message=F, warning=FALSE}
options(mc.cores = parallel::detectCores(), scipen = 0)
#install.packages('rstan')
#install.packages('brms')
library(brms)
library(tidybayes)
library(tidyverse)
```

```{r, echo=F}
sas_nogrid <- sas_nogrid <- theme(panel.background = element_rect(fill = 'white', colour = 'black'), 
                    panel.border = element_rect(color='black', size=1.2, fill = '#00000000'), 
                    panel.grid = element_blank(), 
                    text = element_text(family = 'mono', size = 14),
                    axis.ticks.length = unit(0.15, 'inch'), 
                    axis.ticks = element_line(size=0.7), 
                    strip.background = element_rect(fill = 'black', color = 'black'), 
                    strip.text = element_text(color='white', face = 'bold'))

```



For this demonstration, I will analyze the `palmerpenguins` dataset made available on github by [Allison Horst](https://allisonhorst.github.io/palmerpenguins/articles/intro.html#exploring-factors-1). These data describe multiple species of penguins -- Chinstrap, Gentoo, and Adelie -- that inhabit Antarctica. 

```{r, message=F, warning=F}
#remotes::install_github('allisonhorst/palmerpenguins')
library(palmerpenguins)
data('penguins')
d <- penguins
head(d)
```
## The most simple model

The values that I want to predict in this data set are body mass, measured in grams (`body_mass_g`). The simplest model we could use to predict body mass would contain a single parameter, the `Intercept`, sometimes also called $\alpha$ in statistical notation. This model estimates a grand mean body mass of all penguins in the data set, and the uncertainty around the estimate. We specify this model in `brm` using the formula: `body_mass_g ~ 1`. 

```{r, warning=F, message=F, eval=F}
mod0 <- brm(data = d, family = gaussian(),
            body_mass_g ~ 1)
```

```{r, warning=F, message=F,echo=F}
mod0 <- brm(data = d, family = gaussian(),
            body_mass_g ~ 1, 
            file = "~/Shane's Projects/folio/content/post/2021-12-10-building-multilevel-models-with-brms/models/Model0-Intercept",
            file_refit = 'on_change')
```

We can examine the estimates using `posterior_summary()` -- which returns the estimates for all estimated parameters in a `<brmsfit>` object. The model estimates mean body mass (g) to be `4200.43` with `44.09` grams of error. This comparable to the value of ``r round(mean(d$body_mass_g, na.rm=T), digits=2)`` computed using `base` `R`. The Gaussian family also estimates a `sigma` value of `802.90`, which is comparable to the standard deviation (sd  = ``r round(sd(d$body_mass_g, na.rm=T),digits=2)``)

```{r}
posterior_summary(mod0)
```

Another way to comprehend the mean and uncertainty is to generate draws from the posterior and graph the distribution of body mass values. 

```{r, echo=F, fig.width=4, fig.height=3, fig.cap='Posterior distribution of body mass (g).', fig.align='center'}
mod0 %>% spread_draws(b_Intercept) %>%
    ggplot(aes(b_Intercept)) + geom_density(fill='#3300ff33', color='#3300ff', lwd=1.2) + sas_nogrid + geom_vline(xintercept = 4200.43, lty=2)
```


## Group variation




