---
title: 'Unit 1: Foundational concepts in network science'
author: Shane A. Scaggs
date: '2021-12-12'
slug: unit-1-foundational-network-concepts
categories:
  - Networks
tags:
  - social network analysis
  - teaching
  - concepts
  - theory
  - instruction
draft: yes
---

```{r, message=F, warning=F, echo=F}
library(plotly)
library(tidyverse)
sas_nogrid <- sas_nogrid <- theme(panel.background = element_rect(fill = 'white', colour = 'black'), 
                                  panel.border = element_rect(color='black', size=1.2, fill = '#00000000'), 
                                  panel.grid = element_blank(), 
                                  text = element_text(family = 'mono', size = 14, color = 'black'),
                                  axis.ticks.length = unit(0.15, 'inch'), 
                                  axis.ticks = element_line(size=0.7), 
                                  strip.background = element_rect(fill = 'black', color = 'black'), 
                                  strip.text = element_text(color='white', face = 'bold'))

```

Since the first published use of the term "network analysis" in 1930 by W.H. Martin, published research articles using this phrase have grown in number exponentially. 

```{r, echo=F, fig.cap='Number of research articles on Web of Science containing the phrase "network analysis."'}
netcite <- read.csv("~/Shane's Projects/folio/content/networks/2021-12-12-unit-1-network-theory-and-science/files/network citations.csv")

colnames(netcite) <- c('Year','Citations')
d <- netcite[-c(1,75),]
d$Year <- as.numeric(d$Year)
d$First <- ''

d[ d$Year == 1930, 3] <- 'W.H. Martin. 1930.\nTransmitted frequency range for message circuits.\nBell System Technical Journal.'

plot <- d %>% ggplot(aes(Year, Citations)) +
  geom_line(color='#3300ff66') + 
  geom_point(color='#3300ff') + 
  sas_nogrid + 
  geom_area(fill='#3300ff33', alpha=0.2) + 
  scale_x_continuous(breaks = seq(1930,2020,30)) + 
  scale_y_continuous(breaks = seq(0,15000,3000)) + labs(y=NULL)
ggplotly(plot) 
```

 The incredible growth and spread of network analysis is due to it's flexibility both a method of analysis *and* as a metaphor for the phenomena that researchers study. Network analysis is mathematically rooted in graph theory [cite] and has been applied to problems in diverse academic disciplines including anthropology and sociology, biology and ecology, physics and engineering. The network concept has been used to describe patterns of voting [cite], microbiological interactions [cite], human social relationships [cite], interspecies competition and mutualism [cite], utility grids [cite], the size of the internet and social media [cite], and language structures [cite]. These examples hardly scratch the surface. 

## What is a network? 

How do we define what a *network* is? And what makes it such a widely applicable concept? 