---
title: "Curriculum Vitae"
author: "Shane A. Scaggs"
#date: "7/22/2021"
#output: html_document
---

```{r setup, include=FALSE, message=F, warning=F}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(tidyverse)
```

## Shane A. Scaggs
<p style="text-align: left;">
Graduate Research Fellow<br>
[Human Complexity Lab](https://u.osu.edu/hclab/)<br>
Department of Anthropology<br>
The Ohio State University<br>
scaggs.32@osu.edu</p>

Download a copy of my full CV [here](static/files/CV.2.pdf)

### Research Interests 

Human Subsistence $\times$ Cooperation and Interspecies Coexistence $\times$ Dietary Diversity $\times$ Food Webs $\times$ Network Science $\times$ Complex Social-Ecological Systems $\times$ Community and Wildlife Ecology  

### Education

```{r,echo=F}
edu <- read.csv(file = "~/Shane's Projects/folio/static/csv/edu.csv")
edu %>% kbl(align = 'l') %>% column_spec(1, width = '0.5in')
```

### Publications

```{r,echo=F}
pubs <- openxlsx::read.xlsx("~/Shane's Projects/folio/static/csv/pubs.xlsx", sheet=1)
colnames(pubs) <- c('Year','Citation','DOI','Type')

pubs$Citation <- paste(pubs$Citation, pubs$DOI, sep = ' ')

pubs[,1:2] %>% kbl(align = 'l') %>%
    column_spec(1, width = '0.5in')

```

#### Contributions

```{r, echo=F, message=F, warning=F, fig.width=8, fig.height=4, fig.cap="Contribution matrix.", fig.align='center'}
cont <- openxlsx::read.xlsx("~/Shane's Projects/folio/static/csv/pubs.xlsx", sheet=2)

cont %>% 
  gather(key=key, value=value, -Contribution) %>%
  mutate(value = ifelse(is.na(value)==T, 'No', 'Yes')) %>%
  ggplot() + geom_tile(aes(x=key, y=Contribution, fill=factor(value)), color='black',size=1) + 
  scale_fill_manual(values = c('white','magenta')) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 40, hjust = 1), 
        axis.text = element_text(color='black'),
        text = element_text(size=14, color='black')) + 
  labs(x=NULL, y=NULL, fill='Contributed?') 
```


### Funding Support

```{r,echo=F}
funds <- read.csv(file = "~/Shane's Projects/folio/static/csv/funds.csv")
colnames(funds) <- c("Year","Title","Funding Agency","Amount")
funds %>% kbl(align = 'l') %>% column_spec(1, width = '0.5in')
```

### Conference Proceedings 

```{r,echo=F}
conf <- read.csv(file = "~/Shane's Projects/folio/static/csv/conf.csv")
colnames(conf) <- c("Year","Title","Meeting","Type")
conf[,1:4] %>% 
    arrange(desc( Year)) %>%
    kbl(align = 'l') %>% 
    column_spec(1, width = '0.5in')

```
